Carrier phase and amplitude estimation using pilots and data

### 2012-05-06 paper.tex
Added more to paper.

@@

Again let $a = \rho e^{j\theta}$ and write the sum of squares function  
\[
SS(\rho, \theta, \{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{y_i}^2 - \rho y_i e^{-j\theta} s_i^* - \rho y_i e^{-\theta} s_i + \rho^2.
\]
Differentiating with respect to $\rho$ and setting the resulting expression to zero gives 
\[
\rho = \frac{Z + Z^*}{2L}
\]
where $Z = \sum_{i\in P \cup D}y_i e^{-j\theta} s_i^*$.  Observe that $Z$ is really a function of $\theta$ and $\{d_i, i \in D\}$, but we have supressed the arguments for notational clarity.  We can now write
\[
SS(\rho, \theta, \{d_i, i \in D\}) = A - \rho Z - \rho Z^* + L \rho^2
\]
and substituting the above expression for $\rho$ into this gives the sum of squares function conditioned upon minimisation with respect to $\rho$,
\[
SS(\theta, \{d_i, i \in D\}) = A - \frac{1}{L}\abs{Z}^2.
\]
Substituting~(\ref{eq:hatdfinxtheta}) into this gives the sum of squares function conditioned upon minimisation with respect to $\rho$ and the data sequence $\{d_i, i \in D\}$,
\[
SS(\theta) = A - \frac{1}{L}\abs{Z(\theta)}^2
\]
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta).
\]
We have reused $Z$ here, but this should not cause confusion as $Z(\theta)$ and $Z(\theta, \{d_i, i \in D\})$ are easily told apart from there arguments.  From now on, when we right $Z$ without arguments, we will always be refering to $Z(\theta)$.

We are interested in analysing $\hat{\theta}$ the maximiser of $\abs{Z(\theta)}^2$.  Put $\lambda_L = \langle\theta_0 - \hat{\theta}\rangle$.  We want to show that $\lambda_L$ converges to zero as the number of transmitted symbols $L$ increases to $\infty$ where $\sfracpart{x}$ denotes $x$ taken `modulo $2\pi$' into the interval $[-\pi, \pi)$, that is
\[
\fracpart{x} = x - 2\pi\operatorname{round}\left(\frac{x}{2\pi}\right).
\]
Under the assumption that the noise $\{w_i\}$ is circularly symetric then
\[
y_i = a_0 s_i + w_i = \rho_j e^{j ( \theta_0 + \Phi_i + k_i) }
\] 
where $k_i = \angle s_i$ and the $\Phi_i$ are identically distributed.  The $\Phi_i$ can be identified as \emph{circular random variables}~\cite{Mardia_directional_statistics,Fisher1993,McKilliam_mean_dir_est_sq_arc_length2010}.  Recalling the definition of $\hat{d}_i(\theta)$ and $\hat{u}_i(\theta)$ from~\eqref{eq:hatdfinxtheta}, we have
\begin{align*}
\hat{u}_i(\theta) &= \round{\angle{y_i} - \theta} \\
&= \round{\theta_0 + \Phi_i + k_i - \theta} \\
&\equiv \round{ \fracpart{\theta_0 - \theta} + \Phi_i + k_i} \pmod 2\pi \\
&= \round{ \lambda + \Phi_i + k_i } \\
\end{align*}
where $\lambda = \fracpart{\theta_0 - \theta}$.  So,
\begin{align*}
 y_i e^{-j\theta} \hat{d}_i(\theta)^* &= \rho_i e^{j(\lambda + \Phi_i + k_i - \round{\lambda + \Phi_i + k_i})} \\
&= \rho_i e^{j(\lambda + \Phi_i - \round{\lambda + \Phi_i})}
\end{align*}
when $i \in D$ and
\[
y_i e^{-j\theta} p_i^* = \rho_i e^{j(\lambda + \Phi_i)}
\]
when $i \in P$.  Now we can write
\[
Z(\theta) = \sum_{i \in P} \rho_i e^{j(\lambda + \Phi_i)} + \sum_{i \in D}  \rho_i e^{j( \lambda + \Phi_i - \round{\lambda + \Phi_i})}.
\]
Put $S_L(\lambda) = \frac{1}{L}\abs{Z(\theta)}^2$ and put $\lambda_L = \langle\theta_0 - \hat{\theta}\rangle$.  Then $\lambda_L$ is the maximiser of $S_L(\lambda)$.  We will show that $\lambda_L$ converges almost surely to zero as $L \rightarrow \infty$.  From this, the proof of strong consistency follows.

@@

\appendix

\section{Matched filtering: samples to symbols}

The continuous time transmitted signal takes the form,
\[
x(t) = \sum_{i \in \ints} s_i g(t - iT),
\]
where $g(t)$ is the transmission pulse shape, $T$ is the symbol period.  Assume that the received signal undergoes time and phase offset, is attenuated by some amount and is also subject to additive noise.  The model for the analogue received signal is,
\begin{align*}
r(t) &=  w(t) + \alpha_0 x(t - \tau_0) \\
&= v(t) + \sum_{i \in \ints} s_i g(t - iT - \tau_0),
\end{align*}
where $\alpha_0 \in \complex$ is a complex number representing both attenuation and phase offset, $\tau_0$ is the time offset and $w(t)$ is a continuous noise process.  Assume that an accurate estimate of the time offset $\tau_0$ is available.  We would like to process the signal $r(t)$ so as to obtain a signal in the form of~\eqref{eq:sigmod}.  This is the task of \emph{matched filtering}.

Consider the inner product
\begin{align*}
\dotprod{r(t)}{g(T_sn - Ti -  \tau_0)} &= \dotprod{v(t) + \sum_{k \in \ints} s_k g(t - kT - \tau_0)}{g(T_sn - Ti - \tau)}  \\
&= w_i +  \dotprod{\alpha_0 \sum_{k \in \ints} s_k g(t - kT - \tau_0)}{g(T_sn - Ti - \tau)} \\
&= w_i +  \alpha_0 \sum_{k \in \ints} s_k \dotprod{ g(t - kT - \tau_0)}{g(T_sn - Ti - \tau)} \\
&= w_i + \alpha_0s_i = y_i
\end{align*}
where we put $w_i = \dotprod{v(t)}{g(T_sn - Ti - \tau)}$, and we have required that the transmission pulse $g(t)$ and it's shifts $g(t - Ti)$ are orthogonal, i.e.
\begin{equation}
\dotprod{g(t - iT)}{g(t - kT)} = \int_{\infty}^{\infty} g(t - iT)g^*(t-kT) \, dt = \begin{cases} 1 & i=k \\ 0 & i \neq k, \end{cases} \label{eq:contorthpulse}
\end{equation}
So, the recieved symbols can be obtained by computing the inner product of the recieved sequence signal with the pulse.

Thus far, we have assumed continuous processing of a continuous recieved signal.  However, typically in modern communications, only a sampled version of $r(t)$ is available at the reciever, i.e.
\[
r_n = r(nT_s) = v_n + \alpha_0 x(T_sn - \tau_0),
\] 
where $n$  is an integer and $T_s$ is the sampling period and the sequence $\{v_n, n \in \ints \}$ is a discrete noise process.  In this case, we replace the continuous time inner product, with the discrete time inner product, i.e. put
\[
\dotprod{r(t)}{g(t - iT - \tau)} = \sum_{n\in\ints}r(nT_s)g^*(T_sn - Ti - \tau_0) = \sum_{n\in\ints} r_n g^*(T_sn - Ti - \tau_0)
\]
The inner product above is an infinite sum and probably cannot be computed in practice.  However, in practical systems the number of samples observed is finite.  In what follows we assume that $N$ consecutive samples $r_1, r_2, \dots, r_N$ are observed and we set $r_n = 0$ whenever $n \notin \{1, 2, \dots, N\}$.  Now, the inner products are always finite sums
\[
\dotprod{r(t)}{g(t - iT - \tau_0)} =  \sum_{n=1}^N r_n g^*(T_sn - Ti - \tau_0).
\]
These finite inner products are computable in practice.  Computing the inner products directly using the summation above is not necessarily the most efficient approach.  Instead, the fast Fourier transfrom can be used, as we now show.

We assume that $T_s$ is rationally related to $T$, i.e. $T_s = \frac{p}{q}T$ for some relatively prime integers $p$ and $q$.  This property is a common assumption in the literature.  In fact, it is common to assume the stronger condition, that $T$ is a multiple of $T_s$~\cite{Oerder_synch_square_circstat_2008}.  At the very least, this assumption can be made an accurate approximation by an appropriate choice of $p$ and $q$~\cite{Mordell_Diophantine_equations_1969,Clarkson_thesis}.

Let 
\[
g_n = g^*(-\tfrac{T_s}{p} n - \tau_0).
\]
Then
\[
g_{k-n} = g^*(-\tfrac{T_s}{p}(k - n) - \tau_0) = g^*(n\tfrac{T_s}{p} - k\tfrac{T_s}{p} - \tau_0) = g^*(n\tfrac{T_s}{p} - \tfrac{k}{q}T - \tau_0)
\]
Let
\begin{equation} \label{eq:zzerofill}
z_n = \begin{cases} 
r_{n/c} & n/p \in \{1, \dots, N \} \\
0 & \text{otherwise}
\end{cases}
\end{equation}
be a zero filled version of $r_n$.  Now
\begin{align*}
h_k &= \dotprod{r(t)}{g(t - \tfrac{k}{q}T - \tau_0)} \\
&=  \sum_{n\in\ints} r_n g^*(T_sn - \tfrac{k}{q}T - \tau_0) \\
&= \sum_{n\in\ints} z_n g^*(\tfrac{T_s}{p}n - \tfrac{k}{q}T - \tau_0) \\
&= \sum_{n\in\ints} z_n g_{k-n}.
\end{align*}
So $h_k$ is the convolution between the sequence $\{z_n\}$ and the sequence $\{g_n\}$.  The convoultion can be computed using the fast Fourier transform.  The length of the transform required is
\[
BLERG.
\]  
It only remains to observe that $y_i = h_{iq}$.

>>>>

\begin{theorem}
Let $\{w_i\}$ be a sequence independent and identically distributed, zero mean, circularly symetric complex random variables and let $\{\Phi_i\}$ be the sequence of idenpendent and identically distributed random variables given by $\Phi_i = \angle{(\rho_0 + w_i)}$.  Put 
\[
h_1 = \prob e^{j\round{\Phi_1}} \qquad \text{and} \qquad h_2 = \prob w_1 e^{j\round{\Phi_1}}.
\]
Given the observed noisy data symbols $\{y_i, i \in P \cup D\}$ from~\eqref{eq:sigmod} let $\hat{a}$ be the least squares estimator of $a_0$.  Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}$ and $a_0 = \rho_0 e^{j\theta_0}$.  Put $L = \abs{P \cup D}$.  If 
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d
\] 
as $L \rightarrow \infty$ for positive real numbers $p$ and $d$, and if both $h_1$ and $h_2$ are positive real numbers, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}} \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0( p + d h_1 ) + h_2$ almost surely as $L \rightarrow \infty$,
\item Put $\lambda_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $m_L = \hat{\rho} - \rho_0( p + d h_1 ) + h_2$.  Then the vector $[\lambda_L, m_L]$ converges to the bivariate normal distribution with zero mean and covariance
\[
C = ?
\]
as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

Again let $a = \rho e^{j\theta}$ and write the sum of squares function  
\[
SS(\rho, \theta, \{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{y_i}^2 - \rho y_i e^{-j\theta} s_i^* - \rho y_i e^{-\theta} s_i + \rho^2.
\]
Differentiating with respect to $\rho$ and setting the resulting expression to zero gives 
\[
\rho = \frac{Z + Z^*}{2L}
\]
where $Z = \sum_{i\in P \cup D}y_i e^{-j\theta} s_i^*$.  Observe that $Z$ is really a function of $\theta$ and $\{d_i, i \in D\}$, but we have supressed the arguments for notational clarity.  We can now write
\[
SS(\rho, \theta, \{d_i, i \in D\}) = A - \rho Z - \rho Z^* + L \rho^2
\]
and substituting the above expression for $\rho$ into this gives the sum of squares function conditioned upon minimisation with respect to $\rho$,
\[
SS(\theta, \{d_i, i \in D\}) = A - \frac{1}{L}\abs{Z}^2.
\]
Substituting~(\ref{eq:hatdfinxtheta}) into this gives the sum of squares function conditioned upon minimisation with respect to $\rho$ and the data sequence $\{d_i, i \in D\}$,
\[
SS(\theta) = A - \frac{1}{L}\abs{Z(\theta)}^2
\]
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta).
\]
We have reused $Z$ here, but this should not cause confusion as $Z(\theta)$ and $Z(\theta, \{d_i, i \in D\})$ are easily told apart from there arguments.  %From now on, when we write $Z$ without arguments, we will always be refering to $Z(\theta)$.

We are interested in analysing $\hat{\theta}$ the maximiser of $\abs{Z(\theta)}^2$.  Put $\lambda_L = \langle\theta_0 - \hat{\theta}\rangle$.  We want to show that $\lambda_L$ converges to zero as the number of transmitted symbols $L$ increases to $\infty$ where $\sfracpart{x}$ denotes $x$ taken `modulo $2\pi$' into the interval $[-\pi, \pi)$, that is
\[
\fracpart{x} = x - 2\pi\operatorname{round}\left(\frac{x}{2\pi}\right).
\]
Under the assumption that the noise $\{w_i\}$ is circularly symetric then
\[
y_i = a_0 s_i + w_i = \rho_j e^{j ( \theta_0 + \Phi_i + k_i) }
\] 
where $k_i = \angle s_i$ and the $\Phi_i$ are identically distributed.  The $\Phi_i$ can be identified as \emph{circular random variables}~\cite{Mardia_directional_statistics,Fisher1993,McKilliam_mean_dir_est_sq_arc_length2010}.  Recalling the definition of $\hat{d}_i(\theta)$ and $\hat{u}_i(\theta)$ from~\eqref{eq:hatdfinxtheta}, we have
\begin{align*}
\hat{u}_i(\theta) &= \round{\angle{y_i} - \theta} \\
&= \round{\theta_0 + \Phi_i + k_i - \theta} \\
&\equiv \round{ \fracpart{\theta_0 - \theta} + \Phi_i + k_i} \pmod 2\pi \\
&= \round{ \lambda + \Phi_i + k_i },
\end{align*}
where $\lambda = \fracpart{\theta_0 - \theta}$.  So,
\begin{align*}
 y_i e^{-j\theta} \hat{d}_i(\theta)^* &= \rho_i e^{j(\lambda + \Phi_i + k_i - \round{\lambda + \Phi_i + k_i})} \\
&= \rho_i e^{j(\lambda + \Phi_i - \round{\lambda + \Phi_i})}
\end{align*}
when $i \in D$ and
\[
y_i e^{-j\theta} p_i^* = \rho_i e^{j(\lambda + \Phi_i)}
\]
when $i \in P$.  Now we can write
\[
Z(\theta) = \sum_{i \in P} \rho_i e^{j(\lambda + \Phi_i)} + \sum_{i \in D}  \rho_i e^{j( \lambda + \Phi_i - \round{\lambda + \Phi_i})}.
\]
Put $S_L(\lambda) = \frac{1}{L}\abs{Z(\theta)}^2$ and put $\lambda_L = \langle\theta_0 - \hat{\theta}\rangle$.  Then $\lambda_L$ is the maximiser of $S_L(\lambda)$.  We will show that $\lambda_L$ converges almost surely to zero as $L \rightarrow \infty$.  From this, the proof of strong consistency follows.

@@
